Modelling the Retinotectal projection - can competition provide a stopping mechanism?

### 2021-11-16 paper/paper.tex
The first results section on the chemo-only model

Because there is convincing evidence that EphA and EphB receptors are
expressed in exponentially increasing
patterns \citep{reber_relative_2004,feldheim_genetic_2000,brown_topographic_2000,koulakov_stochastic_2004},
we use an exponential form for retinal receptor expressions, in common with
other modelling
studies~\citep{reber_relative_2004,koulakov_stochastic_2004,simpson_simple_2011}.
We adopt the same precise form for the retinal receptor expression
as \citet{simpson_simple_2011} which is
\begin{equation}
r_i(x,y) = 1.05 + 0.26 \exp(2.3 x)
\end{equation}

The simulated tectum expresses the same types of receptors and ligands, also
in orthogonal pairs of gradients. In this work, we did not consider the effect
of reverse signalling from tectal receptors to RGC ligands so we only modelled
tectal ligand expression, $L_i$. Although several studies model tectal ligand
expression with exponential functions \citep{koulakov_stochastic_2004}, the
experimental evidence for ligand expression is more ambiguous (Supp. Fig
\textbf{X}). Correspondingly, we leave open the possibility that tectal ligand
expression may be modelled by exponential, linear or logarithmic functions:
\begin{equation} \label{e:tecexp}
L_i^{\text{exp}}(x,y) = 1.05 + 0.26 \exp(2.3 x)
\end{equation}
or
\begin{equation}
L_i^{\text{lin}}(x,y) = 1.31 + 2.33 x;
\end{equation}
or
\begin{equation}
L_i^{\text{log}}(x,y) = 2.32 + 1.29 \log (2.3(x+0.2))
\end{equation}
Each function has approximately the same expression values at $x=0$ and
$x=1$. Opposing functions are obtained by substituting $(1-x)$ for $x$.

@@

% All the extra bits of info required to specify the simulation
To enable a simulation of the model described above it is necessary to make
some assumptions about the nature of the receptor-ligand signalled
interactions (i.e. the values of $F_i$ that determine whether interactions are
attractive or repulsive) and about the pattern of ligand expression on the
tectal surface. Unless stated otherwise, we assumed that \textbf{i)} All
receptor-ligand signalled interactions are repulsive, as for EphA-ephrinA
coupling \citep{drescher_vitro_1995,nakamoto_topographically_1996};
and \textbf{ii)} all tectal ligand expressions vary exponentially along a
given axis, following Eq.\,\ref{e:tecexp}.

@@

The first question we tried to answer was whether chemoaffinity alone,
implemented as gradient following, is sufficient to pattern the tectum. We set
$m_1=0.002$ and $m_2=m_3=m_4=0$ and ran the model for 300 steps.

% ./build/sim/agent/agent1 configs/a1/m_ee_G.json  configs/a1/e_wt.json 300
\begin{figure}
\includegraphics[width=\linewidth]{./images/j4_ee_G_wt.png}
\caption{The pure gradient-following model.}
\label{f:chemo}
\end{figure}

@@

To enable a simulation it is necessary to make assumptions about the nature of
the receptor-ligand signalled interactions (i.e. whether they are attractive
or repulsive) and about the pattern of ligand expression on the tectal
surface. We assume that i) All receptor-ligand signalled interactions are
repulsive, as for EphA-ephrinA
coupling \citep{drescher_vitro_1995,nakamoto_topographically_1996}; and ii)
all tectal ligand expressions vary exponentially along a given axis, following
Eq.\,\ref{e:tecexp}. The results of running such a system are presented in
Fig.\,\ref{f:wt}.
%
Fig.\,\ref{f:wt}Ai shows a dual-colour map of retinal ganglion cell positions
on the square retina, with red encoding position on the $x$ axis (temporal to
nasal) and green encoding position on the $y$ axis (ventral to
dorsal). Fig.\,\ref{f:wt}Aii presents a fishnet plot of the expected location on
the tectal surface of each RGC at the end of the development process. Thus,
the `red' RGC from the ventro-nasal corner should map to the caudo-medial
corner of the tectum and the `green' RGC from the dorso-temporal corner should
find its way to the rostro-lateral corner on the tectum. The lines between the
dots of Fig.\,\ref{f:wt}Aii indicate the adjacency relationships between
cells. Thus, colour in each panel indicates a branch or axon's location of
origin on the retina.
%
Fig.\,\ref{f:wt}B shows the initial state of the system, with branches
stochastically arranged around the rostral edge of the tectum. Each dot
represents the centroid of $n=8$ branches per RGC axon. Figs.\,\ref{f:wt}C and \ref{f:wt}D
show the development of the system with time, with Fig.\,\ref{f:wt}D showing a
well ordered arrangement, with a majority of adjacency relationships matching
the experimental prediction (i.e. there are few crossed lines). The
intermediate timepoint at $t=40$ indicates that branches quickly cover the simulated
tectal surface, with final adjacency relationships developing over a longer
timescale. This dynamic behaviour is also evident in Fig.\,\ref{f:wt}E, which
shows \textbf{RMS error or some metric} of the axon centroids with respect to
the experimental prediction. Fig.\,\ref{f:wt}F shows the final
location of each branch in the simulation, which gives a more continuous view
of the order present in Fig.\,\ref{f:wt}D. Finally, Fig.\,\ref{f:wt}G shows the
branches and their centroid path history for five selected RGCs, with path
histories strongly resembling those presented by \citet{simpson_simple_2011}.

@@

We now describe the behaviour of a model which has only the gradient-based
chemoaffinity effect and no axon-axon interaction or border effect
(i.e.~$m_2=m_3=0$). This model was run with $n=8$ branches per RGC axon. All
receptor-ligand signals induced repulsive interactions (i.e.~gradient
descent). Receptor and ligand expressions were as shown in Fig.\,\ref{f:ex},
noting that tectal ligand expression was of exponential form
(Eq.\,\ref{e:tecexp}). Fig.\,\ref{f:ch}A gives a colour map of retinal
ganglion cell soma locations, with the colour red encoding $x$ position on the
temporal-nasal axis (with red signifying nasal positions). Green encodes $y$
position on the dorso-ventral axis.
%
Fig.\,\ref{f:ch}B shows the expected final location of each RGC axon,
according to experimental observations, in a `fishnet' plot which draws lines
between adjacent axons giving a visual indication of the topology of the
arrangement.
%
Fig.\,\ref{f:ch}C shows the position of
400 branches at $t=500$ (in arbitrary time units). Considering each
coordinate axis separately, each branch interacts differentially with two
opposing gradients. Branches which settle nearest the rostral edge of the
tectum interact more strongly with the gradient $L_0$ than with $L_2$
(Fig\,\ref{f:ex}C). A similar pair of gradients separate branches along the
M-L tectal axis. The branches become positioned on a grid due to the
discretization of the tectal ligand gradient (as the number of branches and
tectal elements tends towards infinity, the arangement tends towards a
continuous map---see Fig.\,S1 \textbf{FIXME}).

>>>>

Because there is convincing evidence that EphA and EphB receptors are
expressed in exponentially increasing
patterns \citep{reber_relative_2004,feldheim_genetic_2000,brown_topographic_2000,koulakov_stochastic_2004},
we use an exponential form for retinal receptor expressions, in common with
other modelling
studies~\citep{reber_relative_2004,koulakov_stochastic_2004,simpson_simple_2011}.
We adopt the same precise form for the retinal receptor expression
as \citet{simpson_simple_2011} which is
\begin{equation} \label{e:retrcpt}
r_i(x,y) = 1.05 + 0.26 \exp(2.3 x)
\end{equation}

The simulated tectum expresses the same types of receptors and ligands, also
in orthogonal pairs of gradients. In this work, we did not consider the effect
of reverse signalling from tectal receptors to RGC ligands so we only modelled
tectal ligand expression, $L_i$. Although several studies model tectal ligand
expression with exponential functions \citep{koulakov_stochastic_2004}, the
experimental evidence for ligand expression is more ambiguous (Supp. Fig
\textbf{X}). Correspondingly, we leave open the possibility that tectal ligand
expression may be modelled by exponential, linear or logarithmic functions:
\begin{equation} \label{e:tecligexp}
L_i^{\text{exp}}(x,y) = 1.05 + 0.26 \exp(2.3 x)
\end{equation}
or
\begin{equation}
L_i^{\text{lin}}(x,y) = 1.31 + 2.33 x;
\end{equation}
or
\begin{equation}
L_i^{\text{log}}(x,y) = 2.32 + 1.29 \log (2.3(x+0.2))
\end{equation}
Each function has approximately the same expression values at $x=0$ and
$x=1$. Opposing functions are obtained by substituting $(1-x)$ for $x$.

@@

% All the extra bits of info required to specify the simulation
To enable a simulation of the model described above it is necessary to make
some assumptions about the nature of the receptor-ligand signalled
interactions (i.e. the values of $F_i$ that determine whether interactions are
attractive or repulsive) and about the pattern of ligand expression on the
tectal surface. Unless stated otherwise, we assumed that \textbf{i)} All
receptor-ligand signalled interactions are repulsive, as for EphA-ephrinA
coupling \citep{drescher_vitro_1995,nakamoto_topographically_1996};
and \textbf{ii)} all tectal ligand expressions vary exponentially along a
given axis, following Eq.\,\ref{e:tecligexp}.

@@

% ./build/sim/agent/agent1 configs/simpler/m_ee_G.json  configs/simpler/e_wt_fig1.json
\begin{figure}
\includegraphics[width=\linewidth]{./images/j4_ee_G_wt_fig2.png}
\caption{The pure gradient-following model.}
\label{f:chemo}
\end{figure}

The first question we tried to answer was whether chemoaffinity alone,
implemented as gradient following, is sufficient to pattern the tectum. We set
$m_1=0.002$ and $m_2=m_3=m_4=0$ and ran the model for 300
steps. Fig.\,\ref{f:chemo}A indicates the colour code used for retinal cell
position; dorsal retinal cells are indicated by green, nasal cells by
red. Lightness gives position along the relevant axis so that cells at the
dorsal-nasal corner are yellow and those at the ventral-transverse corner are
black. Fig.\,\ref{f:chemo}B shows the expected layout of retinal ganglion cell
axons on the tectum. The axons retain the colour of their originating cell on
the retina, as shown in Fig.\,\ref{f:chemo}A. Nasal cell axons should be
arranged caudally; dorsal axons laterally; axons should be evenly spaced and
the pattern should span the entire tectum. Figs.\,\ref{f:chemo}C
and \ref{f:chemo}D show, respectively, the positions of branches (i.e.~growth
cones) and axon centroids at $t=300$. Topographic ordering is
correct. Opposing gradients have sorted the axons along the two axes to form a
grid of cells. However, the arrangement does not span the full
tectum. Fig.\,\ref{f:chemo}E shows selected axons (one from the centre and one
from each corner of the retina) and their position history.
Fig.\,\ref{f:chemo}F shows two pattern metrics; the sum of squared distance
errors of the axon centroids and the number of crossings in the 'fish net'
plot of Fig.\,\ref{f:chemo}D which gives an indication of the disorder in the
arrangement. The expected number of crossings is 0, as in
Fig.\,\ref{f:chemo}B.

The form of the retinal receptor expression (Eq.\,\ref{e:retrcpt}) is based on
empirical measurement, but the tectal ligand expression
(Eq.\,\ref{e:tecligexp}) was chosen arbitrarily. To expand the pattern in
Fig.\,\ref{f:chemo}D so that it covers the entire tectum, we needed only to
adjust the tectal ligand expression, substituting the alternative exponential
expression:
%
\begin{equation} \label{e:tecligexp2}
L_i^{\text{exp}}(x,y) = 1.05 + 0.26 \exp(1.2 x)
\end{equation}

% ./build/sim/agent/agent1 configs/simpler/m_eE_G.json  configs/simpler/e_wt_fig3.json
\begin{figure}
\includegraphics[width=\linewidth]{./images/j4_eE_G_wt_fig2.png}
\caption{The pure gradient-following model.}
\label{f:chemo}
\end{figure}

@@

To enable a simulation it is necessary to make assumptions about the nature of
the receptor-ligand signalled interactions (i.e. whether they are attractive
or repulsive) and about the pattern of ligand expression on the tectal
surface. We assume that i) All receptor-ligand signalled interactions are
repulsive, as for EphA-ephrinA
coupling \citep{drescher_vitro_1995,nakamoto_topographically_1996}; and ii)
all tectal ligand expressions vary exponentially along a given axis, following
Eq.\,\ref{e:tecligexp}. The results of running such a system are presented in
Fig.\,\ref{f:wt}.
%
Fig.\,\ref{f:wt}Ai shows a dual-colour map of retinal ganglion cell positions
on the square retina, with red encoding position on the $x$ axis (temporal to
nasal) and green encoding position on the $y$ axis (ventral to
dorsal). Fig.\,\ref{f:wt}Aii presents a fishnet plot of the expected location on
the tectal surface of each RGC at the end of the development process. Thus,
the `red' RGC from the ventro-nasal corner should map to the caudo-medial
corner of the tectum and the `green' RGC from the dorso-temporal corner should
find its way to the rostro-lateral corner on the tectum. The lines between the
dots of Fig.\,\ref{f:wt}Aii indicate the adjacency relationships between
cells. Thus, colour in each panel indicates a branch or axon's location of
origin on the retina.
%
Fig.\,\ref{f:wt}B shows the initial state of the system, with branches
stochastically arranged around the rostral edge of the tectum. Each dot
represents the centroid of $n=8$ branches per RGC axon. Figs.\,\ref{f:wt}C and \ref{f:wt}D
show the development of the system with time, with Fig.\,\ref{f:wt}D showing a
well ordered arrangement, with a majority of adjacency relationships matching
the experimental prediction (i.e. there are few crossed lines). The
intermediate timepoint at $t=40$ indicates that branches quickly cover the simulated
tectal surface, with final adjacency relationships developing over a longer
timescale. This dynamic behaviour is also evident in Fig.\,\ref{f:wt}E, which
shows \textbf{RMS error or some metric} of the axon centroids with respect to
the experimental prediction. Fig.\,\ref{f:wt}F shows the final
location of each branch in the simulation, which gives a more continuous view
of the order present in Fig.\,\ref{f:wt}D. Finally, Fig.\,\ref{f:wt}G shows the
branches and their centroid path history for five selected RGCs, with path
histories strongly resembling those presented by \citet{simpson_simple_2011}.

@@

We now describe the behaviour of a model which has only the gradient-based
chemoaffinity effect and no axon-axon interaction or border effect
(i.e.~$m_2=m_3=0$). This model was run with $n=8$ branches per RGC axon. All
receptor-ligand signals induced repulsive interactions (i.e.~gradient
descent). Receptor and ligand expressions were as shown in Fig.\,\ref{f:ex},
noting that tectal ligand expression was of exponential form
(Eq.\,\ref{e:tecligexp}). Fig.\,\ref{f:ch}A gives a colour map of retinal
ganglion cell soma locations, with the colour red encoding $x$ position on the
temporal-nasal axis (with red signifying nasal positions). Green encodes $y$
position on the dorso-ventral axis.
%
Fig.\,\ref{f:ch}B shows the expected final location of each RGC axon,
according to experimental observations, in a `fishnet' plot which draws lines
between adjacent axons giving a visual indication of the topology of the
arrangement.
%
Fig.\,\ref{f:ch}C shows the position of
400 branches at $t=500$ (in arbitrary time units). Considering each
coordinate axis separately, each branch interacts differentially with two
opposing gradients. Branches which settle nearest the rostral edge of the
tectum interact more strongly with the gradient $L_0$ than with $L_2$
(Fig\,\ref{f:ex}C). A similar pair of gradients separate branches along the
M-L tectal axis. The branches become positioned on a grid due to the
discretization of the tectal ligand gradient (as the number of branches and
tectal elements tends towards infinity, the arangement tends towards a
continuous map---see Fig.\,S1 \textbf{FIXME}).
